The Sound of Teaching Music: Expert Pianists' Performance Modulations for Novices

### 2021-07-11 paper/manuscript.Rmd
remove latex/pdf
Signed-off-by: Atsuko Tominaga <Tominaga_Atsuko@phd.ceu.edu>

bibliography      : ["references.bib"]

floatsintext      : yes
figurelist        : no
tablelist         : no
footnotelist      : no
linenumbers       : no
mask              : no
draft             : no

documentclass     : "apa6"
classoption       : "man"
output: papaja::apa6_word

---

@@

# 1. Introduction

@@

# 2. Experiment 1

## 2.1. Method
```{r methods-1, include = FALSE}
source(here("experiment-1/analysis/demographics", "questionnaire.R"), chdir = TRUE)
source(here("experiment-1/analysis/preprocessor", "error_summary.R"), chdir = TRUE)
```

### 2.1.1. Participants

@@

### 2.1.2. Apparatus and stimuli

@@

One musical excerpt was used as a stimulus. The excerpt was taken from ``A Dozen a Day - Play with Ease in Many Keys" by Edna-Mae Burnam and modified for the experiment. It consisted of a 6-measure isochronous melody noted in 4/4 meter. The stimulus was composed in C major to be played with the right hand only. Non-expressive sheet music (i.e., sheet music without expressive notations, *Figure \ref{fig:stimuli}, A*) was used for the purpose of practice. Expressive notations were added to the non-expressive sheet music for the experiment. They referred to either articulation or dynamics (*Figure \ref{fig:stimuli}, B-C*). Articulation was notated as either legato or staccato. Legato indicates that musical notes are to be connected and should sound smooth. Staccato requires producing musical notes with shortened duration, keeping them separate from each other. Dynamics was notated as either forte and piano. Forte indicates that musical notes should be played loudly whereas piano indicates that musical notes should be played softly. The notation did not include any indication of fingering (i.e., the positioning of the fingers when playing the piano) because the piece was simple and pilot testing had shown that specifying fingering was not necessary.

### 2.1.3. Procedure

First, participants were allocated to either the teaching or performing condition and asked to practise the excerpt with the notated expression of either articulation or dynamics (*Figure \ref{fig:stimuli}, B-C*). As soon as they had produced the excerpt with the notated expression without pitch errors twice consecutively, the test trials began.

@@

### 2.1.4. Data analysis

@@

```{r stim-1, out.width = "100%", fig.cap = "\\label{fig:stimuli}(A)Non-expressive sheet music. (B)Articulation. The curved line (slur) indicates legato and the dots indicate staccato. (C)Dynamics. The symbol `f' denotes forte and the symbol `p' denotes piano. For data analysis, only the 16th notes were included.", echo = FALSE}
stim_a <- image_border(image_read(here("paper/image/sheetmusic-1", "stim_a.png")), "#FFFFFF", "120x120")
stim_d <- image_border(image_read(here("paper/image/sheetmusic-1", "stim_d.png")), "#FFFFFF", "120x120")
stim_n <- image_border(image_read(here("paper/image/sheetmusic-1", "stim_n.png")), "#FFFFFF", "120x120")

# combine
img_stimuli <- image_append(c(stim_n, stim_a, stim_d), stack = TRUE)
img_stimuli <- image_annotate(img_stimuli, "(A)", size = 80, location = "+100+0", color = "black", font = "helvetica")
img_stimuli <- image_annotate(img_stimuli, "(B)", size = 80, location = "+100+650", color = "black", font = "helvetica")
img_stimuli <- image_annotate(img_stimuli, "(C)", size = 80, location = "+100+1300", color = "black", font = "helvetica")
img_stimuli
```

@@

## 2.2. Results

@@

All effects are reported as significant at *p* < .05. We first report results for performance of the piece with the notated expression of articulation (*Figure \ref{fig:stimuli}, B*), followed by performance of the piece with the notated expression of dynamics (*Figure \ref{fig:stimuli}, C*). To recall our predictions, if participants play more slowly when they are trying to teach, interonset intervals (IOIs) should be larger when teaching. If participants specifically modulate relevant aspects of the expressive techniques they are trying to teach, key-overlap time (KOT) should reveal exaggerated legato/staccato when teaching articulation, and key velocity (KV) should reveal amplifications of forte and piano when teaching dynamics.

### 2.2.1. Articulation: Interonset Intervals (IOIs)

In order to compare the mean IOIs between the teaching and performing condition, we conducted a Wiscoxon Signed-rank test, instead of a paired *t*-test, because a Shapiro-Wilk test showed that the distribution of the mean difference was significantly different from the normal distribution (*p* `r if(ioi_art_norm$p.value < 0.001){"< 0.001"} else {paste("= ", sprintf('%.3f', ioi_art_norm$p.value))}`). The Wiscoxon Signed-rank test revealed that participants played more slowly in the teaching condition [*Mdn* = `r ioi_art_all[Condition == "teaching"]$Median` (ms), *IQR* = `r ioi_art_all[Condition == "teaching"]$IQR`] than in the performing condition [*Mdn* = `r ioi_art_all[Condition == "performing"]$Median` (ms), *IQR* = `r ioi_art_all[Condition == "performing"]$IQR`] while playing the piece with the notated articulation (*p* `r if(ioi_art_wilcoxon$p.value < 0.001){"< 0.001"} else {paste("= ", sprintf('%.3f', ioi_art_wilcoxon$p.value))}`, *r* = `r round(wilcoxeff_art$effsize, 2)`, *Figure \ref{fig:ioi-1}*).

@@

### 2.2.2. Articulation: Key-Overlap Time (KOT)

A two-way repeated-measures ANOVA with the factors Condition (teaching vs. performing) and Subcomponent-LS (legato vs. staccato) revealed that there was a significant main effect of Subcomponent-LS (*F*(`r kot_art_aov$anova_table[2,1]`,`r kot_art_aov$anova_table[2,2]`) = `r sprintf('%.0f', kot_art_aov$anova_table[2,4])`, *p* `r if(kot_art_aov$anova_table[2,6] < 0.001){"< 0.001"} else {paste("= ", sprintf('%.3f', kot_art_aov$anova_table[2,6]))}`, $\eta_G^2$ = `r sprintf('%.2f', kot_art_aov$anova_table[2,5])`) and a significant interaction between Condition and Subcomponent-LS (*F*(`r kot_art_aov$anova_table[3,1]`,`r kot_art_aov$anova_table[3,2]`) = `r sprintf('%.2f', kot_art_aov$anova_table[3,4])`, *p* `r if(kot_art_aov$anova_table[3,6] < 0.001){"< 0.001"} else {paste("= ", sprintf('%.3f', kot_art_aov$anova_table[3,6]))}`, $\eta_G^2$ = `r sprintf('%.3f', kot_art_aov$anova_table[3,5])`, *Figure \ref{fig:kot-1}*). Post-hoc comparisons based on the estimated marginal means with Tukey adjustment showed that participants produced shorter staccato in the teaching condition [*M* = `r kot_art_all[Condition == "teaching" & Subcomponent == "Staccato"]$Mean` (ms), *SD* = `r kot_art_all[Condition == "teaching" & Subcomponent == "Staccato"]$SD`] than in the performing condition [*M* = `r kot_art_all[Condition == "performing" & Subcomponent == "Staccato"]$Mean` (ms), *SD* = `r kot_art_all[Condition == "performing" & Subcomponent == "Staccato"]$SD`] (*p* = `r sprintf('%.3f', kot_art_posthoc$p.value[2])`) while there was no significant difference in legato between the teaching [*M* = `r kot_art_all[Condition == "teaching" & Subcomponent == "Legato"]$Mean` (ms), *SD* = `r kot_art_all[Condition == "teaching" & Subcomponent == "Legato"]$SD`] and performing condition [*M* = `r kot_art_all[Condition == "performing" & Subcomponent == "Legato"]$Mean` (ms), *SD* = `r kot_art_all[Condition == "performing" & Subcomponent == "Legato"]$SD`] (*p* = `r sprintf('%.2f', kot_art_posthoc$p.value[1])`).

@@

### 2.2.3. Articulation: Key Velocity (KV)

A two-way repeated-measures ANOVA with the factors Condition (teaching vs. performing) and Subcomponent-LS (legato vs. staccato) showed that there was a significant interaction between Condition and Subcomponent-LS (*F*(`r vel_art_aov$anova_table[3,1]`,`r vel_art_aov$anova_table[3,2]`) = `r sprintf('%.2f', vel_art_aov$anova_table[3,4])`, *p* `r if(vel_art_aov$anova_table[3,6] < 0.001){"< 0.001"} else {paste("= ", sprintf('%.3f', vel_art_aov$anova_table[3,6]))}`, $\eta_G^2$ = `r sprintf('%.3f', vel_art_aov$anova_table[3,5])`, *Figure \ref{fig:vel-1}*). However, post-hoc comparisons based on the estimated marginal means with Tukey adjustment did not find a significant difference between the teaching [Legato: *M* = `r vel_art_all[Condition == "teaching" & Subcomponent == "Legato"]$Mean`, *SD* = `r vel_art_all[Condition == "teaching" & Subcomponent == "Legato"]$SD`; Staccato: *M* = `r vel_art_all[Condition == "teaching" & Subcomponent == "Staccato"]$Mean`, *SD* = `r vel_art_all[Condition == "teaching" & Subcomponent == "Staccato"]$SD`] and performing condition [Legato: *M* = `r vel_art_all[Condition == "performing" & Subcomponent == "Legato"]$Mean`, *SD* = `r vel_art_all[Condition == "performing" & Subcomponent == "Legato"]$SD`, Staccato: *M* = `r vel_art_all[Condition == "performing" & Subcomponent == "Staccato"]$Mean`, *SD* = `r vel_art_all[Condition == "performing" & Subcomponent == "Staccato"]$SD`] for each subcomponent (Legato: *p* = `r sprintf('%.2f', vel_art_posthoc$p.value[1])`, Staccato: *p* = `r sprintf('%.2f', vel_art_posthoc$p.value[2])`).

@@

### 2.2.4. Dynamics: Interonset Intervals (IOIs)

In order to compare the mean IOIs between the teaching and performing condition, we conducted a Wiscoxon Signed-rank test, instead of a paired *t*-test, because a Shapiro-Wilk test showed that the distribution of the mean difference was significantly different from the normal distribution (*p* `r if(ioi_dyn_norm$p.value < 0.001){"< 0.001"} else {paste("= ", sprintf('%.3f', ioi_dyn_norm$p.value))}`). The Wiscoxon Signed-rank test revealed no significant difference between the teaching condition [*Mdn* = `r ioi_dyn_all[Condition == "teaching"]$Median` (ms), *IQR* = `r ioi_dyn_all[Condition == "teaching"]$IQR`] and the performing condition [*Mdn* = `r ioi_dyn_all[Condition == "performing"]$Median` (ms), *IQR* = `r ioi_dyn_all[Condition == "performing"]$IQR`] while playing the piece with the notated dynamics (*p* `r if(ioi_dyn_wilcoxon$p.value < 0.001){"< 0.001"} else {paste("= ", sprintf('%.2f', ioi_dyn_wilcoxon$p.value))}`, *r* = `r round(wilcoxeff_dyn$effsize, 2)`, *Figure \ref{fig:ioi-1}*).

@@

### 2.2.5. Dynamics: Key Velocity (KV)

A two-way repeated-measures ANOVA with the factors Condition (teaching vs. performing) and Subcomponent-FP (forte vs. piano) showed that there was a significant main effect of Subcomponent-FP (*F*(`r vel_dyn_aov$anova_table[2,1]`,`r vel_dyn_aov$anova_table[2,2]`) = `r sprintf('%.0f', vel_dyn_aov$anova_table[2,4])`, *p* `r if(vel_dyn_aov$anova_table[2,6] < 0.001){"< 0.001"} else {paste("= ", sprintf('%.3f', vel_dyn_aov$anova_table[2,6]))}`, $\eta_G^2$ = `r sprintf('%.2f', vel_dyn_aov$anova_table[2,5])`) and a significant interaction between Condition and Subcomponent-FP (*F*(`r vel_dyn_aov$anova_table[3,1]`,`r vel_dyn_aov$anova_table[3,2]`) = `r sprintf('%.2f', vel_dyn_aov$anova_table[3,4])`, *p* `r if(vel_dyn_aov$anova_table[3,6] < 0.001){"< 0.001"} else {paste("= ", sprintf('%.3f', vel_dyn_aov$anova_table[3,6]))}`, $\eta_G^2$ = `r sprintf('%.3f', vel_dyn_aov$anova_table[3,5])`, *Figure \ref{fig:vel-1}*). Post-hoc comparisons based on the estimated marginal means with Tukey adjustment revealed that participants produced louder forte in the teaching condition [*M* = `r vel_dyn_all[Condition == "teaching" & Subcomponent == "Forte"]$Mean`, *SD* = `r vel_dyn_all[Condition == "teaching" & Subcomponent == "Forte"]$SD`] than in the performing condition [*M* = `r vel_dyn_all[Condition == "performing" & Subcomponent == "Forte"]$Mean`, *SD* = `r vel_dyn_all[Condition == "performing" & Subcomponent == "Forte"]$SD`] (*p* `r if(vel_dyn_posthoc$p.value[1] < 0.001){"< 0.001"} else {paste("= ", sprintf('%.3f', vel_dyn_posthoc$p.value[1]))}`) while there was no significant difference in piano between the teaching [*M* = `r vel_dyn_all[Condition == "teaching" & Subcomponent == "Piano"]$Mean`, *SD* = `r vel_dyn_all[Condition == "teaching" & Subcomponent == "Piano"]$SD`] and performing condition [*M* = `r vel_dyn_all[Condition == "performing" & Subcomponent == "Piano"]$Mean`, *SD* = `r vel_dyn_all[Condition == "performing" & Subcomponent == "Piano"]$SD`] (*p* `r if(vel_dyn_posthoc$p.value[2] < 0.001){"< 0.001"} else {paste("= ", sprintf('%.2f', vel_dyn_posthoc$p.value[2]))}`).

@@

### 2.2.6. Dynamics: KV Transition Points

Additionally, we conducted an exploratory analysis by focusing on specific points at which each subcomponent changed from one to the other (i.e., forte to piano: FtoP, piano to forte: PtoF). These points could be structurally important in order to make a contrast between forte and piano. We calculated the KV difference for each interval by subtracting the KV value of the current note from that of the following note. Outliers were removed using the same criteria as for the other dependent variables and less than 5 % of the data were excluded from analysis. A two-way repeated-measures ANOVA with the factors Condition (teaching vs. performing) and Subcomponent-FP (FtoP vs. PtoF) showed that there was a significant main effect of Subcomponent-FP (*F*(`r vel_diff_dyn_aov$anova_table[2,1]`,`r vel_diff_dyn_aov$anova_table[2,2]`) = `r sprintf('%.0f', vel_diff_dyn_aov$anova_table[2,4])`, *p* `r if(vel_diff_dyn_aov$anova_table[2,6] < 0.001){"< 0.001"} else {paste("= ", sprintf('%.3f', vel_diff_dyn_aov$anova_table[2,6]))}`, $\eta_G^2$ = `r sprintf('%.2f', vel_diff_dyn_aov$anova_table[2,5])`) and a significant interaction between Condition and Subcomponent-FP (*F*(`r vel_diff_dyn_aov$anova_table[3,1]`,`r vel_diff_dyn_aov$anova_table[3,2]`) = `r sprintf('%.2f', vel_diff_dyn_aov$anova_table[3,4])`, *p* `r if(vel_diff_dyn_aov$anova_table[3,6] < 0.001){"< 0.001"} else {paste("= ", sprintf('%.3f', vel_diff_dyn_aov$anova_table[3,6]))}`, $\eta_G^2$ = `r sprintf('%.3f', vel_diff_dyn_aov$anova_table[3,5])`, *Figure \ref{fig:vel-diff}*). Post-hoc comparisons based on the estimated marginal means with Tukey adjustment showed that there was a larger KV difference when the expressive notation changed from forte to piano (FtoP: *p* `r if(vel_diff_dyn_posthoc$p.value[1] < 0.001){"< 0.001"} else {paste("= ", sprintf('%.3f', vel_diff_dyn_posthoc$p.value[1]))}`) and from piano to forte (PtoF: *p* `r if(vel_diff_dyn_posthoc$p.value[2] < 0.001){"< 0.001"} else {paste("= ", sprintf('%.3f', vel_diff_dyn_posthoc$p.value[2]))}`) in the teaching condition [FtoP: *M* = `r vel_diff_dyn_all[Condition == "teaching" & Subcomponent == "FtoP"]$Mean`, *SD* = `r vel_diff_dyn_all[Condition == "teaching" & Subcomponent == "FtoP"]$SD`; PtoF: *M* = `r vel_diff_dyn_all[Condition == "teaching" & Subcomponent == "PtoF"]$Mean`, *SD* = `r vel_diff_dyn_all[Condition == "teaching" & Subcomponent == "PtoF"]$SD`] than in the performing condition [FtoP: *M* = `r vel_diff_dyn_all[Condition == "performing" & Subcomponent == "FtoP"]$Mean`, *SD* = `r vel_diff_dyn_all[Condition == "performing" & Subcomponent == "FtoP"]$SD`, PtoF: *M* = `r vel_diff_dyn_all[Condition == "performing" & Subcomponent == "PtoF"]$Mean`, *SD* = `r vel_diff_dyn_all[Condition == "performing" & Subcomponent == "PtoF"]$SD`].

@@

### 2.2.7. Dynamics: Key-Overlap Time (KOT)

>>>>

bibliography      : ["reference.bib"]
csl: plos.csl

floatsintext      : no
figurelist        : no
tablelist         : no
footnotelist      : no
linenumbers       : yes
mask              : no
draft             : no

output: papaja::apa6_word
---

@@

# Introduction

@@

# Experiment 1

## Materials and Methods
```{r methods-1, include = FALSE}
source(here("experiment-1/analysis/demographics", "questionnaire.R"), chdir = TRUE)
source(here("experiment-1/analysis/preprocessor", "error_summary.R"), chdir = TRUE)
```

### Participants

@@

### Apparatus and stimuli

@@

One musical excerpt was used as a stimulus. The excerpt was taken from ``A Dozen a Day - Play with Ease in Many Keys" by Edna-Mae Burnam and modified for the experiment. It consisted of a 6-measure isochronous melody noted in 4/4 meter. The stimulus was composed in C major to be played with the right hand only. Non-expressive sheet music (i.e., sheet music without expressive notations, *Fig \ref{fig:stimuli}, A*) was used for the purpose of practice. Expressive notations were added to the non-expressive sheet music for the experiment. They referred to either articulation or dynamics (*Fig \ref{fig:stimuli}, B-C*). Articulation was notated as either legato or staccato. Legato indicates that musical notes are to be connected and should sound smooth. Staccato requires producing musical notes with shortened duration, keeping them separate from each other. Dynamics was notated as either forte and piano. Forte indicates that musical notes should be played loudly whereas piano indicates that musical notes should be played softly. The notation did not include any indication of fingering (i.e., the positioning of the fingers when playing the piano) because the piece was simple and pilot testing had shown that specifying fingering was not necessary.

### Procedure

First, participants were allocated to either the teaching or performing condition and asked to practise the excerpt with the notated expression of either articulation or dynamics (*Fig \ref{fig:stimuli}, B-C*). As soon as they had produced the excerpt with the notated expression without pitch errors twice consecutively, the test trials began.

@@

### Data analysis

@@

```{r stim-1, out.width = "100%", fig.cap = "\\label{fig:stimuli}(A)Non-expressive sheet music. (B)Articulation. The curved line (slur) indicates legato and the dots indicate staccato. (C)Dynamics. The symbol `f' denotes forte and the symbol `p' denotes piano. For data analysis, only the 16th notes were included.", echo = FALSE}
stim_a <- image_border(image_read(here("paper/sheetmusic/1", "stim_a.png")), "#FFFFFF", "120x120")
stim_d <- image_border(image_read(here("paper/sheetmusic/1", "stim_d.png")), "#FFFFFF", "120x120")
stim_n <- image_border(image_read(here("paper/sheetmusic/1", "stim_n.png")), "#FFFFFF", "120x120")

# combine
img_stimuli <- image_append(c(stim_n, stim_a, stim_d), stack = TRUE)
img_stimuli <- image_annotate(img_stimuli, "(A)", size = 80, location = "+100+0", color = "black", font = "helvetica")
img_stimuli <- image_annotate(img_stimuli, "(B)", size = 80, location = "+100+650", color = "black", font = "helvetica")
img_stimuli <- image_annotate(img_stimuli, "(C)", size = 80, location = "+100+1300", color = "black", font = "helvetica")
img_stimuli

# save
image_write(img_stimuli, path = here("paper/image/", "stimuli.tiff"), format = "tiff")
```

@@

## Results

@@

All effects are reported as significant at *p* < .05. We first report results for performance of the piece with the notated expression of articulation (*Fig \ref{fig:stimuli}, B*), followed by performance of the piece with the notated expression of dynamics (*Fig \ref{fig:stimuli}, C*). To recall our predictions, if participants play more slowly when they are trying to teach, interonset intervals (IOIs) should be larger when teaching. If participants specifically modulate relevant aspects of the expressive techniques they are trying to teach, key-overlap time (KOT) should reveal exaggerated legato/staccato when teaching articulation, and key velocity (KV) should reveal amplifications of forte and piano when teaching dynamics.

## Articulation
### Interonset Intervals (IOIs)

In order to compare the mean IOIs between the teaching and performing condition, we conducted a Wiscoxon Signed-rank test, instead of a paired *t*-test, because a Shapiro-Wilk test showed that the distribution of the mean difference was significantly different from the normal distribution (*p* `r if(ioi_art_norm$p.value < 0.001){"< 0.001"} else {paste("= ", sprintf('%.3f', ioi_art_norm$p.value))}`). The Wiscoxon Signed-rank test revealed that participants played more slowly in the teaching condition [*Mdn* = `r ioi_art_all[Condition == "teaching"]$Median` (ms), *IQR* = `r ioi_art_all[Condition == "teaching"]$IQR`] than in the performing condition [*Mdn* = `r ioi_art_all[Condition == "performing"]$Median` (ms), *IQR* = `r ioi_art_all[Condition == "performing"]$IQR`] while playing the piece with the notated articulation (*p* `r if(ioi_art_wilcoxon$p.value < 0.001){"< 0.001"} else {paste("= ", sprintf('%.3f', ioi_art_wilcoxon$p.value))}`, *r* = `r round(wilcoxeff_art$effsize, 2)`, *Fig \ref{fig:ioi-1}*).

@@

### Articulation: Key-Overlap Time (KOT)

A two-way repeated-measures ANOVA with the factors Condition (teaching vs. performing) and Subcomponent-LS (legato vs. staccato) revealed that there was a significant main effect of Subcomponent-LS (*F*(`r kot_art_aov$anova_table[2,1]`,`r kot_art_aov$anova_table[2,2]`) = `r sprintf('%.0f', kot_art_aov$anova_table[2,4])`, *p* `r if(kot_art_aov$anova_table[2,6] < 0.001){"< 0.001"} else {paste("= ", sprintf('%.3f', kot_art_aov$anova_table[2,6]))}`, $\eta_G^2$ = `r sprintf('%.2f', kot_art_aov$anova_table[2,5])`) and a significant interaction between Condition and Subcomponent-LS (*F*(`r kot_art_aov$anova_table[3,1]`,`r kot_art_aov$anova_table[3,2]`) = `r sprintf('%.2f', kot_art_aov$anova_table[3,4])`, *p* `r if(kot_art_aov$anova_table[3,6] < 0.001){"< 0.001"} else {paste("= ", sprintf('%.3f', kot_art_aov$anova_table[3,6]))}`, $\eta_G^2$ = `r sprintf('%.3f', kot_art_aov$anova_table[3,5])`, *Fig \ref{fig:kot-1}*). Post-hoc comparisons based on the estimated marginal means with Tukey adjustment showed that participants produced shorter staccato in the teaching condition [*M* = `r kot_art_all[Condition == "teaching" & Subcomponent == "Staccato"]$Mean` (ms), *SD* = `r kot_art_all[Condition == "teaching" & Subcomponent == "Staccato"]$SD`] than in the performing condition [*M* = `r kot_art_all[Condition == "performing" & Subcomponent == "Staccato"]$Mean` (ms), *SD* = `r kot_art_all[Condition == "performing" & Subcomponent == "Staccato"]$SD`] (*p* = `r sprintf('%.3f', kot_art_posthoc$p.value[2])`) while there was no significant difference in legato between the teaching [*M* = `r kot_art_all[Condition == "teaching" & Subcomponent == "Legato"]$Mean` (ms), *SD* = `r kot_art_all[Condition == "teaching" & Subcomponent == "Legato"]$SD`] and performing condition [*M* = `r kot_art_all[Condition == "performing" & Subcomponent == "Legato"]$Mean` (ms), *SD* = `r kot_art_all[Condition == "performing" & Subcomponent == "Legato"]$SD`] (*p* = `r sprintf('%.2f', kot_art_posthoc$p.value[1])`).

@@

### Key Velocity (KV)

A two-way repeated-measures ANOVA with the factors Condition (teaching vs. performing) and Subcomponent-LS (legato vs. staccato) showed that there was a significant interaction between Condition and Subcomponent-LS (*F*(`r vel_art_aov$anova_table[3,1]`,`r vel_art_aov$anova_table[3,2]`) = `r sprintf('%.2f', vel_art_aov$anova_table[3,4])`, *p* `r if(vel_art_aov$anova_table[3,6] < 0.001){"< 0.001"} else {paste("= ", sprintf('%.3f', vel_art_aov$anova_table[3,6]))}`, $\eta_G^2$ = `r sprintf('%.3f', vel_art_aov$anova_table[3,5])`, *Fig \ref{fig:vel-1}*). However, post-hoc comparisons based on the estimated marginal means with Tukey adjustment did not find a significant difference between the teaching [Legato: *M* = `r vel_art_all[Condition == "teaching" & Subcomponent == "Legato"]$Mean`, *SD* = `r vel_art_all[Condition == "teaching" & Subcomponent == "Legato"]$SD`; Staccato: *M* = `r vel_art_all[Condition == "teaching" & Subcomponent == "Staccato"]$Mean`, *SD* = `r vel_art_all[Condition == "teaching" & Subcomponent == "Staccato"]$SD`] and performing condition [Legato: *M* = `r vel_art_all[Condition == "performing" & Subcomponent == "Legato"]$Mean`, *SD* = `r vel_art_all[Condition == "performing" & Subcomponent == "Legato"]$SD`, Staccato: *M* = `r vel_art_all[Condition == "performing" & Subcomponent == "Staccato"]$Mean`, *SD* = `r vel_art_all[Condition == "performing" & Subcomponent == "Staccato"]$SD`] for each subcomponent (Legato: *p* = `r sprintf('%.2f', vel_art_posthoc$p.value[1])`, Staccato: *p* = `r sprintf('%.2f', vel_art_posthoc$p.value[2])`).

@@

## Dynamics
### Interonset Intervals (IOIs)

In order to compare the mean IOIs between the teaching and performing condition, we conducted a Wiscoxon Signed-rank test, instead of a paired *t*-test, because a Shapiro-Wilk test showed that the distribution of the mean difference was significantly different from the normal distribution (*p* `r if(ioi_dyn_norm$p.value < 0.001){"< 0.001"} else {paste("= ", sprintf('%.3f', ioi_dyn_norm$p.value))}`). The Wiscoxon Signed-rank test revealed no significant difference between the teaching condition [*Mdn* = `r ioi_dyn_all[Condition == "teaching"]$Median` (ms), *IQR* = `r ioi_dyn_all[Condition == "teaching"]$IQR`] and the performing condition [*Mdn* = `r ioi_dyn_all[Condition == "performing"]$Median` (ms), *IQR* = `r ioi_dyn_all[Condition == "performing"]$IQR`] while playing the piece with the notated dynamics (*p* `r if(ioi_dyn_wilcoxon$p.value < 0.001){"< 0.001"} else {paste("= ", sprintf('%.2f', ioi_dyn_wilcoxon$p.value))}`, *r* = `r round(wilcoxeff_dyn$effsize, 2)`, *Fig \ref{fig:ioi-1}*).

@@

### Key Velocity (KV)

A two-way repeated-measures ANOVA with the factors Condition (teaching vs. performing) and Subcomponent-FP (forte vs. piano) showed that there was a significant main effect of Subcomponent-FP (*F*(`r vel_dyn_aov$anova_table[2,1]`,`r vel_dyn_aov$anova_table[2,2]`) = `r sprintf('%.0f', vel_dyn_aov$anova_table[2,4])`, *p* `r if(vel_dyn_aov$anova_table[2,6] < 0.001){"< 0.001"} else {paste("= ", sprintf('%.3f', vel_dyn_aov$anova_table[2,6]))}`, $\eta_G^2$ = `r sprintf('%.2f', vel_dyn_aov$anova_table[2,5])`) and a significant interaction between Condition and Subcomponent-FP (*F*(`r vel_dyn_aov$anova_table[3,1]`,`r vel_dyn_aov$anova_table[3,2]`) = `r sprintf('%.2f', vel_dyn_aov$anova_table[3,4])`, *p* `r if(vel_dyn_aov$anova_table[3,6] < 0.001){"< 0.001"} else {paste("= ", sprintf('%.3f', vel_dyn_aov$anova_table[3,6]))}`, $\eta_G^2$ = `r sprintf('%.3f', vel_dyn_aov$anova_table[3,5])`, *Fig \ref{fig:vel-1}*). Post-hoc comparisons based on the estimated marginal means with Tukey adjustment revealed that participants produced louder forte in the teaching condition [*M* = `r vel_dyn_all[Condition == "teaching" & Subcomponent == "Forte"]$Mean`, *SD* = `r vel_dyn_all[Condition == "teaching" & Subcomponent == "Forte"]$SD`] than in the performing condition [*M* = `r vel_dyn_all[Condition == "performing" & Subcomponent == "Forte"]$Mean`, *SD* = `r vel_dyn_all[Condition == "performing" & Subcomponent == "Forte"]$SD`] (*p* `r if(vel_dyn_posthoc$p.value[1] < 0.001){"< 0.001"} else {paste("= ", sprintf('%.3f', vel_dyn_posthoc$p.value[1]))}`) while there was no significant difference in piano between the teaching [*M* = `r vel_dyn_all[Condition == "teaching" & Subcomponent == "Piano"]$Mean`, *SD* = `r vel_dyn_all[Condition == "teaching" & Subcomponent == "Piano"]$SD`] and performing condition [*M* = `r vel_dyn_all[Condition == "performing" & Subcomponent == "Piano"]$Mean`, *SD* = `r vel_dyn_all[Condition == "performing" & Subcomponent == "Piano"]$SD`] (*p* `r if(vel_dyn_posthoc$p.value[2] < 0.001){"< 0.001"} else {paste("= ", sprintf('%.2f', vel_dyn_posthoc$p.value[2]))}`).

@@

### KV Transition Points

Additionally, we conducted an exploratory analysis by focusing on specific points at which each subcomponent changed from one to the other (i.e., forte to piano: FtoP, piano to forte: PtoF). These points could be structurally important in order to make a contrast between forte and piano. We calculated the KV difference for each interval by subtracting the KV value of the current note from that of the following note. Outliers were removed using the same criteria as for the other dependent variables and less than 5 % of the data were excluded from analysis. A two-way repeated-measures ANOVA with the factors Condition (teaching vs. performing) and Subcomponent-FP (FtoP vs. PtoF) showed that there was a significant main effect of Subcomponent-FP (*F*(`r vel_diff_dyn_aov$anova_table[2,1]`,`r vel_diff_dyn_aov$anova_table[2,2]`) = `r sprintf('%.0f', vel_diff_dyn_aov$anova_table[2,4])`, *p* `r if(vel_diff_dyn_aov$anova_table[2,6] < 0.001){"< 0.001"} else {paste("= ", sprintf('%.3f', vel_diff_dyn_aov$anova_table[2,6]))}`, $\eta_G^2$ = `r sprintf('%.2f', vel_diff_dyn_aov$anova_table[2,5])`) and a significant interaction between Condition and Subcomponent-FP (*F*(`r vel_diff_dyn_aov$anova_table[3,1]`,`r vel_diff_dyn_aov$anova_table[3,2]`) = `r sprintf('%.2f', vel_diff_dyn_aov$anova_table[3,4])`, *p* `r if(vel_diff_dyn_aov$anova_table[3,6] < 0.001){"< 0.001"} else {paste("= ", sprintf('%.3f', vel_diff_dyn_aov$anova_table[3,6]))}`, $\eta_G^2$ = `r sprintf('%.3f', vel_diff_dyn_aov$anova_table[3,5])`, *Fig \ref{fig:vel-diff}*). Post-hoc comparisons based on the estimated marginal means with Tukey adjustment showed that there was a larger KV difference when the expressive notation changed from forte to piano (FtoP: *p* `r if(vel_diff_dyn_posthoc$p.value[1] < 0.001){"< 0.001"} else {paste("= ", sprintf('%.3f', vel_diff_dyn_posthoc$p.value[1]))}`) and from piano to forte (PtoF: *p* `r if(vel_diff_dyn_posthoc$p.value[2] < 0.001){"< 0.001"} else {paste("= ", sprintf('%.3f', vel_diff_dyn_posthoc$p.value[2]))}`) in the teaching condition [FtoP: *M* = `r vel_diff_dyn_all[Condition == "teaching" & Subcomponent == "FtoP"]$Mean`, *SD* = `r vel_diff_dyn_all[Condition == "teaching" & Subcomponent == "FtoP"]$SD`; PtoF: *M* = `r vel_diff_dyn_all[Condition == "teaching" & Subcomponent == "PtoF"]$Mean`, *SD* = `r vel_diff_dyn_all[Condition == "teaching" & Subcomponent == "PtoF"]$SD`] than in the performing condition [FtoP: *M* = `r vel_diff_dyn_all[Condition == "performing" & Subcomponent == "FtoP"]$Mean`, *SD* = `r vel_diff_dyn_all[Condition == "performing" & Subcomponent == "FtoP"]$SD`, PtoF: *M* = `r vel_diff_dyn_all[Condition == "performing" & Subcomponent == "PtoF"]$Mean`, *SD* = `r vel_diff_dyn_all[Condition == "performing" & Subcomponent == "PtoF"]$SD`].

@@

### Key-Overlap Time (KOT)

A two-way repeated-measures ANOVA with the factors Condition (teaching vs. performing) and Subcomponent-FP (forte vs. piano) revealed that there was a significant main effect of Subcomponent-FP (*F*(`r kot_dyn_aov$anova_table[2,1]`,`r kot_dyn_aov$anova_table[2,2]`) = `r sprintf('%.0f', kot_dyn_aov$anova_table[2,4])`, *p* `r if(kot_dyn_aov$anova_table[2,6] < 0.001){"< 0.001"} else {paste("= ", sprintf('%.3f', kot_dyn_aov$anova_table[2,6]))}`, $\eta_G^2$ = `r sprintf('%.2f', kot_dyn_aov$anova_table[2,5])`, *Fig \ref{fig:kot-1}*), reflecting more key overlap for forte notes compared to piano notes. Neither the main effect of Condition (*F*(`r kot_dyn_aov$anova_table[1,1]`,`r kot_dyn_aov$anova_table[1,2]`) = `r sprintf('%.2f', kot_dyn_aov$anova_table[1,4])`, *p* `r if(kot_dyn_aov$anova_table[1,6] < 0.001){"< 0.001"} else {paste("= ", sprintf('%.2f', kot_dyn_aov$anova_table[1,6]))}`, $\eta_G^2$ = `r sprintf('%.3f', kot_dyn_aov$anova_table[1,5])`) nor the interaction between Condition and Subcomponent-FP (*F*(`r kot_dyn_aov$anova_table[3,1]`,`r kot_dyn_aov$anova_table[3,2]`) = `r sprintf('%.3f', kot_dyn_aov$anova_table[3,4])`, *p* `r if(kot_dyn_aov$anova_table[3,6] < 0.001){"< 0.001"} else {paste("= ", sprintf('%.2f', kot_dyn_aov$anova_table[3,6]))}`, $\eta_G^2$ = `r sprintf('%.3f', kot_dyn_aov$anova_table[3,5])`) was significant.
